---
title: 'How to lie with Statistics: Polish presidential election 2025'
author:
- name: Tomasz Przechlewski
  email: "t.plata-przechlewski@psw.kwidzyn.edu.pl"
  affiliation: Powiślańska Szkoła Wyższa (Kwidzyn/Poland)
date: "27/6/2025"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library("tidyverse")
library("knitr")

```

## Introduction

An explanation of the entirely undeservedly high-profile analysis by
Dr. Eng. Kontek, which purported to scientifically prove mass fraud
committed during the 2025 presidential election in Poland. For reasons
unknown, Mr. Kontek was so convincing that the Prosecutor General,
Mr. Bodnar, took his claims with full seriousness:

*The election protests filed by Krzysztof Kontek and Joanna Staniszkis
were forwarded by the Supreme Court to the Prosecutor General on June
23, 2025, for review and response, with a two-day deadline set for
submitting a position, i.e., by the end of June 25, 2025, in
accordance with the rules for calculating deadlines set out in the
Code of Civil Procedure*.

*The protest authors attached materials indicating possible
irregularities in the vote-counting process. Their allegations were
based on a scientifically verifiable method, which led the Prosecutor
General to conclude that there was a high probability of actual
irregularities in the work of the indicated electoral commissions*.

In this short text, we will explain why Mr. Bodnar has embarrassed
himself and the offices he holds.

## Kontek’s Method

Conceptually, the method is very simple:

1. Kontek assumes that within small geographical areas, voting
preferences are similar (scientifically homogeneous), which means that
support for candidates should be roughly the same across all precincts
in each area.  He identifies 2,200 such areas. On average, each area
contains about 15 precincts and 10,000 voters who cast their ballots
(20 million / 2,200).

2. For each area, he calculates the median and the variability of
support for Karol Nawrocki.  Variability here means how much the
results differ between precincts. To measure this variability, he uses
something called the median absolute deviation (MAD).

3. He then calculates the following measure:

$$k = \frac{x - me}{mad}$$

where $x$ is the support for KN, $me$ is the median support,
and $mad$ is the measure of variability known as MAD.
The formula may look intimidating, but if $(x - me)$ is the
percentage difference between the support in a precinct
and the median for a given group,
then dividing it by $mad$ rescales the result from a percentage
into a multiple of MAD.
So instead of a difference of, say, 4.34%, the result
is 2.3 — meaning the deviation from the median is 2.3 MADs.

He then arbitrarily chooses a threshold value he considers acceptable.
Values with an absolute magnitude greater than 3 are treated as
anomalies.  Kontek uses either $k = 2$ or $k = 3$ as his "acceptable"
thresholds.

Out of around 32,000 precinct-level election commissions, he thus
identified 1,400 anomalies using $k = 3$, or 5,000 anomalies using
$k = 2$

## Why is this analysis flawed?

Item (assumption) 1 — that small geographical areas must exhibit the same
level of support — is pulled out of thin air. He simply assumes it,
but there’s no reason it must be true. His entire “model” is based on
this, and **any model is only as credible
as the assumptions it's built on**.

It’s worth noting that you don’t need any knowledge of statistics to
recognize that his work is worthless garbage — just plain common sense
is enough. And this is what Justice Minister Bodnar took to the
Supreme Court? Honestly, it’s embarrassing.

Items 2--4 are just irrelevant math. Of course his method produces
“anomalies” — it has to. But whether these are signs of fraud is a
completely different question. Kontek simply used an old trick:
mathematically proving something, then pretending that this something
has to do with something else, despite never actually demonstrating
any relationship between the two.

## Example

I had not time to recreate the exact grouping of precincts the way
Kontek did, but the grouping itself shouldn’t matter.
Which specific precincts form a group is irrelevant, as long as they are
geographically adjacent and there are around 15 of them (otherwise,
the whole method is fundamentally flawed).

So, for example, the municipality of Sopot has 25 voting
precincts. (Kontek grouped by 15, but the more precincts in a group,
the harder it is to detect “anomalies” — so my example is more than
valid. The more precincts, the more difficult it is to find
“anomalies”!)

```{r, echo=FALSE}
p25_r2 <- read.csv("protokoly_po_obwodach_w_drugiej_turze_utf8.csv",
                   sep = ';', dec='.',
                   colClasses = c('Teryt.Gminy'= 'character' ),
                   header=TRUE,
                   ##skip=1,
                   na.string="NA" ) |>
  select( nrk=Nr.komisji,
          gmina=Gmina, 
          teryt=Teryt.Gminy, 
          siedziba=Siedziba, 
          lgw=Liczba.głosów.ważnych.oddanych.łącznie.na.obu.kandydatów..z.kart.ważnych.,
          nawrocki=NAWROCKI.Karol.Tadeusz,
          trzaskowski=TRZASKOWSKI.Rafał.Kazimierz ) |>
  ## Poprawiamy teryt
  mutate(teryt = ifelse( nchar(teryt) == 5, sprintf ("0%s", teryt), teryt)) |>
  mutate ( rr =  nawrocki + trzaskowski,
           nawrocki.p = nawrocki/rr *100,
           trzaskowski.p = trzaskowski/rr *100  ) |>
  mutate ( nrk = sprintf("%s.%03i", teryt, nrk)) |>
  select(nrk, teryt, siedziba,
         nawrocki2=nawrocki, trzaskowski2=trzaskowski, 
         nawrocki.p2=nawrocki.p, trzaskowski.p2=trzaskowski.p, lgw2=lgw) |>
  ## kolumna z kodem pocztowym:
  mutate ( pkod = str_extract(siedziba, "\\b\\d{2}-\\d{3}\\b"))

## Tylko Sopot
sopot_r2 <- p25_r2 |>  filter (teryt == '226401')

me <- median(sopot_r2$nawrocki.p2)
mad <- mad(sopot_r2$nawrocki.p2)

sopot_r2_00 <- sopot_r2 |>
  mutate (me = median(nawrocki.p2),
          mad = mad(nawrocki.p2),
          k = (nawrocki.p2 - me) /mad
          ) |>
  #filter (k > 2) |>
  arrange (-k) |>
  select(siedziba, lgw2, `gangus&sutener`=nawrocki.p2, me, mad, k)

```

After applying Kontek’s procedure, we get:
median = `r round(me, 2)`% support for KN (obviously, the elite votes for the right candidate :-))
mad = `r round(mad, 2)`%.

For two precincts, $k > 3$ (top two rows in the table), that’s where
they cheated:

```{r, echo=FALSE}
kable(sopot_r2_00, booktabs = TRUE, digits = 2)
```

The columns `lgw2` and `gangus&sutener` contain, respectively,
the number of valid votes
and the percentage of support for KN.

## Conclusions?

1. The rural rabble came to Zoppot for treatment and lowered Rafał Trzaskowski support (just kidding).
(explanation for foreigners: those 2 anomalities are locate in rehabilitation centers,
   populated for obvious reasons with people not from Sopot)
   
2. In 99% of those 1,400 or 5,000 “anomalous” precincts identified by
   Kontek, the situation is exactly the same as in Sopot.

